---
title: "StructuralVariantAnnotation Quick Overview"
author: "Ruining Dong, Daniel Cameron"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
vignette: >
  %\VignetteIndexEntry{Structural Variant Annotation Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(#echo = TRUE,
  collapse = TRUE,
  comment = "#>")
```

## Introduction
This vignette outlines a workflow of parsing and plotting structural variants from Variant Call Format (VCF) using the `StructuralVariantAnnotation` package. `StructuralVariantAnnotation` contains useful helper functions for reading and interpreting structural variant calls. The package contains functions for parsing VCFs from a number of popular callers, dealing with breakpoints involving two separate genomic loci encoded as `GRanges` objects, as well as identifying various biological phenomena. Currently the identification of retroposed transcripts (RTs) and nuclear mitochondrial integrations (NUMTs) are supported.


## Using GRanges for structural variants: a breakend-centric data structure

This package uses a breakend-centric event notation adopted from the `StructuralVariantAnnotation` package. In short, breakends are stored in a GRanges object with strand used to indicate breakpoint orientation. where breakpoints are represented using a `partner` field containing the name of the breakend at the other side of the breakend. This notation was chosen as it simplifies the annotations of RTs which are detected at breakend-level.

## Workflow
### Loading data from VCF

VCF data is parsed into a `VCF` object using the `readVCF` function from the
Bioconductor package `VariantAnnotation`. Simple filters could be applied to a 
`VCF` object to remove unwanted calls. The `VCF` object is then converted to a `GRanges` object with breakend-centric notations using `StructuralVariantAnnotation`. More information about `VCF` objects and breakend-centric GRanges object can be found by
consulting the vignettes in the corresponding packages with `browseVignettes("VariantAnnotation")` and `browseVignettes("StructuralVariantAnnotation")`.

```{r input, warning=FALSE, message=FALSE}
suppressPackageStartupMessages(require(StructuralVariantAnnotation))
suppressPackageStartupMessages(require(VariantAnnotation))
suppressPackageStartupMessages(require(RTDetect))
vcf.file <- system.file("extdata", "gridss.vcf", package = "RTDetect")
vcf <- VariantAnnotation::readVcf(vcf.file, "hg19")
gr <- StructuralVariantAnnotation::breakpointRanges(vcf)
```

Note that `StructuralVariantAnnotation` requires the `GRanges` object to be composed entirely of valid breakpoints. Please consult the vignette of the `StructuralVariantAnnotation` package for ensuring breakpoint consistency.

### Identifying Retrotransposed Transcripts
The package provides `rtDetect` to identify RTs using the provided SV calls. This is achieved by detecting intronic deletions, which are breakpoints at exon-intron (and intron-exon) boundaries of a transcript. Fusions consisting of an exon boundary and a second genomic location are reported as potential insertion sites. Due to the complexity of RT events, insertion sites can be discovered on both left and right sides, only one side, or none at all.

The output is a list of `GRanges` object consisting of two sets of `GRanges` calls, `insSite` and `junctions`, containing candidate insertion sites and exon-exon junctions respectively. Candidate insertion sites are annotated by the source transcripts and whether exon-exon junctions are detected for the source transcripts. RT junction breakends are annotated by the UCSC exon IDs, corresponding transcripts, and NCBI gene symbols.

```{r}
suppressPackageStartupMessages(require(TxDb.Hsapiens.UCSC.hg19.knownGene))
suppressPackageStartupMessages(require(dplyr))
hg19.genes <- TxDb.Hsapiens.UCSC.hg19.knownGene
RT_vcf <- readVcf(system.file("extdata", "diploidSV.vcf", package = "RTDetect"))
RT_gr <- StructuralVariantAnnotation::breakpointRanges(RT_vcf, nominalPosition=TRUE)
RT <- rtDetect(RT_gr, hg19.genes, maxgap=50, minscore=0.3)
```
```{r}
RT$SKA3
```


## Visualising breakpoint pairs via circos plots

One way of visualising RT is by circos plots. Here we use the package
[`circlize`](https://doi.org/10.1093/bioinformatics/btu393) to demonstrate 
the visualisation of insertion site and exon-exon junctions. 

To generate a simple circos plot of RT event with SKA3 transcript:
```{r}
suppressPackageStartupMessages(require(circlize))
rt_chr_prefix <- c(RT$SKA3$junctions, RT$SKA3$insSite)
seqlevelsStyle(rt_chr_prefix) <- "UCSC"
pairs <- breakpointgr2pairs(rt_chr_prefix)
pairs
```
To see supporting breakpoints clearly, we generate the circos plot according to the loci of event.
```{r}
circos.initializeWithIdeogram(
    data.frame(V1=c("chr13", "chr11"), 
               V2=c(21720000,108585000), 
               V3=c(21755000,108586000), 
               V4=c("q12.11","q24.3"), 
               V5=c("gneg","gpos50")))
circos.genomicLink(as.data.frame(S4Vectors::first(pairs)), as.data.frame(S4Vectors::second(pairs)))
circos.clear()
```

## SessionInfo
```{r}
sessionInfo()
```
